---
title: "DSCI445 Final Presentation"
author: "Kwan Eum, Eddie Haggerty, Nathan Seto"
date: "`r Sys.Date()`"
output: powerpoint_presentation
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, eval = TRUE, warning = FALSE, error = FALSE)
```

# Predicting Problematic Internet Use



## Premise: Kaggle Competition

- Child Mind Institute: Problematic Internet Use
- Main goal of competition is to predict Severity Impairment Index (SII) score
- Also wanted to try to predict Sleep Disturbance score (SDS) out of curiosity

## Exploratory Data Analysis

- “SII” stands for “Severity Impairment Index”

```{r cars, echo = TRUE}
 ```

## Slide with Plot

```{r pressure}
plot(pressure)
```

## Exploratory Data Analysis

“SDS” stands for “Sleep Disturbance Scale”

# Linear Regression


## Dealing with NAs

- Removed variables with proportion of NAs over 0.5
- 10 Variables removed this way
- Then removed all observations with proportion of NAs over 0.5
- 1097 observations removed this way

```{r}
# NA tables
```

## Dealing with NAs

- Used imputation for rest of NAs; mice and tidyr packages were good for this
- There were still some NAs in the data after imputing
- All remaining NAs were in same columns/rows
- Chose to get rid of observations rather than removing more variables
- Final dataset had 61 variables
- Training set had 1584 observations
- Test set had 360 observations

## Lasso

- Trying to predict SDS Score
- 28 predictors were used, including sii
- Only 2 of the 17 Bioelectric Impedance Analysis variables were used
- BMI and Activity Level
- Test RMSE for the best Lasso model was 10.73954

```{r}
# Lasso Plot
```

## Ridge 

- Same 28 predictors were used in both Ridge and Lasso
- Different coefficient estimates
- Test RMSE for best Ridge model was 10.6742

```{r}
# Ridge plot
```

# Random Forest

## Random Forest
Data Cleaning: 
- Removed predictors and observations with more than 40% missing data
- A total of 36 variables and 1,225 observations were excluded
- Applied Random Forest to identify key variables associated with:
1. Severity Impairment Index (SII)
2. Sleep Disturbance Scale (SDS)
- Variable Adjustments
- Excluded the Parent-Child Internet Addiction Test (PCIAT) total score because it is a linear combination of other PCIAT variables.
- PCIAT variables are categorical, ranging from 0 (strongly disagree) to 5 (strongly agree).

## Random Forest

- The plot highlights the most important variables related to the SII.
- PCIAT variables are the most strongly associated with SII.
- Physical measures (Weight, Height, BMI) are the next most important contributors

```{r}
# RF variable importance plot #1
```

## Random Forest

- This plot reflecting their importance in predicting the Sleep Disturbance Scale (SDS).
- PCIAT variables again show the strongest association with SDS.
- Physical measures are the second most important contributors, consistent with the SII results.

```{r}
# RF variable importance plot #2
```

# Linear Discriminant Analysis

## Dealing with NAs

- Options:
1. Imputation 
- Median for numerical variables
- “Missing” for categorical variables
2. Remove rows where response = NA
- Followed with imputation
3. Removing Variables

```{r}
# image
```

## Linear Discriminant Analysis

- Test Data Confusion Matrix: 99.27% accuracy

```{r}
# conf mat
# graph
```

